### 2023-03-21 chapter0/chapter0.tex
Removed reference to regulation which better reflects phrasing in subsequent paragraph
Added paragraphs on IDRs in transcription

The structural diversity of IDRs is directly related to their functional plasticity, and as IDRs are not confined to one conformation, they can interact with and bind to many possible partners. Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions. As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling and regulation. Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life. However, the cells of more complex organisms called eukaryotes contain additional subcompartments called organelles. Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time. However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by the interaction of one protein with another, which in turn creates a change in the state of the system that propagates the signal. Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation. Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein. Phosphate groups contain three negative charges, so their addition can dramatically affect a protein's structural energetics and induce a conformational change. Thus, phosphorylation often plays a key role in toggling proteins between inactive and active states. In general, though, PTMs encompass a variety of chemical modifications with similarly diverse impacts on a protein's behavior. Because IDRs are generally exposed to their environment, they are frequent targets of PTMs. However, PTMs do not occur haphazardly in proteins but are instead targeted to binding sites created by sequential patterns of residues called short linear motifs (SLiMs). SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners. In contrast to the highly structured interfaces that characterize folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}. Thus, their interactions are relatively weak and highly transient. As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously. IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Wright2014}. Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues. Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different genetic ``programs.''

Some of the most prominent examples of IDRs in cell signaling and regulation are the activation domains of transcription factors. Each cell expresses a unique complement of proteins that in large part determines its identity.

% IDRs are enriched in transcriptional machinery
%    A. Introduce transcription via the central dogma
%    B. Describe transcription factors, cofactors, transcriptional machinery
%        1. Examples of different transcription factors
%    C. Describe properties of transcription factors
%        1. DNA binding and activation domains
%        2. Different flavors of activation domains
%    D. How transcription factors find their targets to initiate specific genetic programs is one of the major outstanding questions in molecular biology

% Disorder is spectrum
%    A. MorFs, folding on binding, fuzzy complexes

>>>>

The structural diversity of IDRs is directly related to their functional plasticity, and as IDRs are not confined to one conformation, they can interact with and bind to many possible partners. Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions. As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling. Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life. However, the cells of more complex organisms called eukaryotes contain additional subcompartments called organelles. Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time. However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by the interaction of one protein with another, which in turn creates a change in the state of the system that propagates the signal. Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation. Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein. Phosphate groups contain three negative charges, so their addition can dramatically affect a protein's structural energetics and induce a conformational change. Thus, phosphorylation often plays a key role in toggling proteins between inactive and active states. In general, though, PTMs encompass a variety of chemical modifications with similarly diverse impacts on a protein's behavior. Because IDRs are generally exposed to their environment, they are frequent targets of PTMs. However, PTMs do not occur haphazardly in proteins but are instead targeted to binding sites created by sequential patterns of residues called short linear motifs (SLiMs). SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners. In contrast to the highly structured interfaces that characterize folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}. Thus, their interactions are relatively weak and highly transient. As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously. IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Wright2014}. Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues. Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different genetic ``programs.''

Some of the most prominent examples of IDRs in cell signaling are found in the proteins that regulate the creation of other proteins. A protein's amino acid sequence is encoded in DNA in a unit called a gene, and since all cells in an organism share the same genome, they can in principle synthesize any protein encoded in it.\footnote{The DNA sequences of different cells in an organism are not always strictly identical. For example, spontaneous mutations can create changes that range from single-letter substitutions to large-scale rearrangements. In other cases, such as during the creation of sex or immune cells, a cell's DNA is intentionally modified to generate genetic diversity.} However, different cell types instead express unique complements of proteins that in large part determine their identities. This selective conversion of the information stored in DNA into proteins is called gene expression and is a highly regulated process that varies in space and time, \textit{i.e.} between different cells and within a single cell. Though gene expression involves dozens, if not hundreds, of distinct biochemical reactions, most occur as steps within two major processes: transcription and translation. In transcription, the sequence of a protein encoded the cell's DNA is transcribed into an intermediate molecule called messenger RNA (mRNA). mRNA is also a kind of nucleic acid, and is chemically closely related to DNA. However, as a much smaller molecule containing only the information needed to synthesize a protein, mRNA is more easily transported and manipulated in subsequent steps. Thus, if a gene stored as DNA is a cell's master record of a protein sequence, mRNA is its working copy. In translation, a vast molecular machine called the ribosome then synthesizes a protein by translating the information encoded in a strand of mRNA into a sequence of amino acids.

Both transcription and translation are tightly regulated processes, but because cells employ numerous interlocking mechanisms to control access to DNA and the creation of mRNA, transcription is often the major regulatory checkpoint in gene expression. A key step in this process is the recruitment of the protein complex which transcribes DNA into mRNA to the beginning of a gene's sequence, \textit{i.e.} its transcription start site (TSS). The placement of this complex, called RNA polymerase, is dictated by regulatory sequences encoded in the DNA, and while some regulatory sequences are ubiquitous, like those that mark a gene's TSS, many are gene-specific. Thus, the unique collection of regulatory sequences associated with a gene determines many aspects of its expression. RNA polymerase does not directly contact these gene-specific regulatory elements, however. Instead, proteins called transcription factors directly bind these sequences and interact with other components of the transcriptional machinery which in turn recruit RNA polymerase to the TSS. Accordingly, a typical transcription factor has two parts: a DNA-binding domain and an activation domain. DNA-binding domains are usually structured because they make stable contacts with specific regulatory sequences. In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs. Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences. Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology. As disordered regions play a key role in this process, studies which broadly investigate the relationship between the sequences, structures, and functions of IDRs therefore have significant implications for our understanding of gene regulation.
